%---------------------------------------------------------------------------%
%->> Frontmatter
%---------------------------------------------------------------------------%
%-
%-> 生成封面
%-
\maketitle% 生成中文封面
\MAKETITLE% 生成英文封面
%-
%-> 作者声明
%-
\makedeclaration% 生成声明页
%-
%-> 中文摘要
%-
\intobmk\chapter*{摘\quad 要}% 显示在书签但不显示在目录
\setcounter{page}{1}% 开始页码
\pagenumbering{Roman}% 页码符号

原初黑洞是在宇宙早期由于局域不均匀性导致原初密度扰动坍塌而形成的黑洞。因为原初黑洞不仅可以作为冷暗物质的候选者，而且可以解释LIGO-Virgo科学组织探测到的双黑洞并合事件，所以原初黑洞受到越来越广泛的关注。本文探索通过引力波的手段来探测原初黑洞。

首先，考虑所有其他原初黑洞以及线性密度扰动产生的力矩对原初双黑洞演化的影响，我们计算了原初黑洞具有一般质量谱情况下的原初双黑洞的并合率分布，并证实了绝大多数的冷暗物质不是由恒星级质量的原初黑洞构成的。利用一维和二维的原初黑洞并合率分布，我们有望重构出原初黑洞的质量分布，进而帮助我们理解LIGO-Virgo探测到的双黑洞的形成与演化机制。

其次，我们计算了双黑洞和双中子星并合产生的随机引力波背景。我们考虑了两种不同的双黑洞形成机制，分别是天体物理双黑洞和原初双黑洞机制。我们发现原初双黑洞会比天体物理双黑洞产生更强的随机引力波背景。另外，双黑洞和双中子星产生的随机引力波背景可以被未来的空间引力波探测器LISA探测到。如果这一引力波背景没能从LISA探测器中扣除掉，将会构成LISA的额外噪音，从而降低LISA的探测能力。

然后，我们探讨了通过下一代地基引力波探测器，比如爱因斯坦望远镜和宇宙勘探者，来区分原初黑洞和天体物理黑洞的可能性。通过定向搜寻亚太阳质量的双黑洞系统，我们估算了原初黑洞占冷暗物质丰度的可探测上限。另外，我们预测了爱因斯坦望远镜和宇宙勘探者能够探测到的双黑洞事件数目随红移的分布，从而来区分原初黑洞和天体物理黑洞。

最后，我们在北美纳赫兹引力波天文台11年的数据中搜索伴随原初黑洞形成而产生的标量诱导引力波信号。由于没有发现统计意义上显著的引力波信号，我们对质量在$[2 \times 10^{-3}, 7\times 10^{-1}]$太阳质量区间的原初黑洞占冷暗物质的丰度给出了迄今为止最严格的限制。

\keywords{原初黑洞，引力波，冷暗物质，并合率，随机引力波背景}% 中文关键词


\intobmk\chapter*{Abstract}% 显示在书签但不显示在目录

Primordial black holes can form through the collapse of the primordial density perturbation due to the local inhomogeneity in the early Universe. Primordial black holes have attracted more and more attention as they can not only serve as a candidate of cold dark matter but also explain the binary black holes observed by LIGO-Virgo scientific collaboration. This work is devoted to the investigation of probing the primordial black holes through gravitational waves.

Firstly, we work out the merger rate distribution of primordial-black-hole binaries with a general mass function by taking into account the torques by all primordial black holes and linear density perturbations. We confirm that cold dark matter is not dominated by the stellar-mass primordial black holes. Utilizing the one-dimensional and two-dimensional merger rate distributions, we expect to reconstruct the mass function of primordial black holes, thus helping us to understand the formation and evolution mechanism of the binary black holes observed by LIGO-Virgo. 

Secondly, we calculate the stochastic gravitational-wave background from binary black holes, covering the astrophysical and primordial scenarios separately, together with the one from binary neutron stars. We find primordial-black-hole binaries contribute a stronger stochastic gravitational-wave background than that from astrophysical-black-hole binaries, and the total stochastic gravitational-wave background from both binary black holes and binary neutron stars has a high possibility to be detected by the future space-based detector such as LISA. If not been subtracted, the stochastic gravitational-wave background can contribute an additional source of confusion noise to LISA's total noise curve, thus weakening LISA's detection abilities. 

Thirdly, we investigate how the next generation of ground-based gravitational-wave detectors, such as Einstein Telescope and Cosmic Explorer, can be used to distinguish primordial black holes from astrophysical black holes. We estimate the detectable limits of the abundance of primordial black holes in the cold dark matter by the targeted search for sub-solar mass primordial black hole binaries. We forecast the detectable event rate distributions for the primordial-black-hole binaries and astrophysical-black-hole binaries by Einstein Telescope and Cosmic Explorer, which can serve as a method to distinguish primordial black holes from astrophysical black holes.

Lastly, we perform the direct search for the signals of scalar-induced gravitational waves accompanying the formation of primordial black holes in the North American Nanohertz Observatory for Gravitational waves 11-year data set. No statistically significant detection has been made, and hence we place the most stringent upper limit on the abundance of primordial black holes in the mass range of $[2 \times 10^{-3}, 7\times 10^{-1}] \Msun$.

\KEYWORDS{Primordial Black Holes, Gravitational Waves, Cold Dark Matter, Merger Rate, Stochastic Gravitational-Wave Background}% 英文关键词
%---------------------------------------------------------------------------%
